Maximum Entropy Spectral Analysis: a case study

### 2023-07-12 paper/mesa.tex
added the figures and more text

\subsection{Marginalising over the noise PSD in GW parameter estimation: the case for GW150914}
As an application of MESA, and its implementation in \texttt{memspectrum}, we analyse GW150914\cite{} using a Bayesian framework that allows for the marginalisation of the order  $p$ of the AR($p$) process representing the noise data stream. The inference algorithm proceeds as usual, see \cite{}[veitch and vecchio, ashton et al, etc etc], with a modification to the likelihood calculation. The likelihood function we employ is given in Eq.~\eqref{eq:Whittle} and its calculation proceeds as follows: for all proposed Monte Carlo moves $\theta\equiv\theta_{GW}\cup p$\footnote{We indicate the set of all GW parameters (component masses, spins, luminosity distance, etc.) with $\theta_{GW}$.}, for each detector included in the analysis, we compute the detector response to the template $h(t;\theta_{GW})$ in the time-domain, subtract it from the data $d(t)$ to get the time domain residuals $r(t;\theta_{GW})$. Under the MESA assumption, we describe the PSD of the residuals as in Eq.~\eqref{eq:MESApsd}, where the order $p$ is fixed by the Monte Carlo sampler. Having computed the PSD, we Fourier transform the residuals and compute the likelihood. Our sampler is a nested sampling algorithm\cite{}[raynest] and the specific inference model is implemented as part of \texttt{granite}\cite{}[in preparation], a dedicated inference model for ground based interferometric detectors. Needless to say, despite the implementation of the fast Burg algorithm in \texttt{memspectrum}, the inference with noise marginalisation is vastly more computational expensive than the standard analysis. As our analysis template, we adopt the machine learning based MLGW model\cite{}[MLGW], an aligned spin model trained on TEOBResumS\cite{}, that has been shown to perform well on LVK detected events\cite{}[MLGW]. 
\begin{figure}
	\caption{In the top panel, we show the difference between the forecasted strain values (CAT model) and the actual values observed at Livingston interferometer (blue). The red shadow denotes the $90\%$ confidence interval subtracted with the actual time series values. The confidence interval is computed on  $100$ independent random realization of the predictions.
	The bottom panel displays the standard deviation of the predictions with both CAT and FPE models. The inset shows a detail of the first $\SI{0.5}{s}$ of plot in the top panel.
	}
	\label{fig:gw150914_intrinsic}
	\includegraphics{Images/gw150914/intrinsic.pdf}
\end{figure}
\begin{figure}
	\caption{In the top panel, we show the difference between the forecasted strain values (CAT model) and the actual values observed at Livingston interferometer (blue). The red shadow denotes the $90\%$ confidence interval subtracted with the actual time series values. The confidence interval is computed on  $100$ independent random realization of the predictions.
	The bottom panel displays the standard deviation of the predictions with both CAT and FPE models. The inset shows a detail of the first $\SI{0.5}{s}$ of plot in the top panel.
	}
	\label{fig:gw150914_intrinsic}
	\includegraphics{Images/gw150914/extrinsic.pdf}
\end{figure}
\begin{figure}
	\caption{In the top panel, we show the difference between the forecasted strain values (CAT model) and the actual values observed at Livingston interferometer (blue). The red shadow denotes the $90\%$ confidence interval subtracted with the actual time series values. The confidence interval is computed on  $100$ independent random realization of the predictions.
	The bottom panel displays the standard deviation of the predictions with both CAT and FPE models. The inset shows a detail of the first $\SI{0.5}{s}$ of plot in the top panel.
	}
	\label{fig:gw150914_intrinsic}
	\includegraphics{Images/gw150914/waveforms.pdf}
\end{figure}
\begin{figure}
	\caption{In the top panel, we show the difference between the forecasted strain values (CAT model) and the actual values observed at Livingston interferometer (blue). The red shadow denotes the $90\%$ confidence interval subtracted with the actual time series values. The confidence interval is computed on  $100$ independent random realization of the predictions.
	The bottom panel displays the standard deviation of the predictions with both CAT and FPE models. The inset shows a detail of the first $\SI{0.5}{s}$ of plot in the top panel.
	}
	\label{fig:gw150914_intrinsic}
	\includegraphics{Images/gw150914/posterior_order_H1.pdf}
\end{figure}

\begin{figure}
	\caption{In the top panel, we show the difference between the forecasted strain values (CAT model) and the actual values observed at Livingston interferometer (blue). The red shadow denotes the $90\%$ confidence interval subtracted with the actual time series values. The confidence interval is computed on  $100$ independent random realization of the predictions.
	The bottom panel displays the standard deviation of the predictions with both CAT and FPE models. The inset shows a detail of the first $\SI{0.5}{s}$ of plot in the top panel.
	}
	\label{fig:gw150914_intrinsic}
	\includegraphics{Images/gw150914/posterior_order_L1.pdf}
\end{figure}

\begin{figure}
	\caption{In the top panel, we show the difference between the forecasted strain values (CAT model) and the actual values observed at Livingston interferometer (blue). The red shadow denotes the $90\%$ confidence interval subtracted with the actual time series values. The confidence interval is computed on  $100$ independent random realization of the predictions.
	The bottom panel displays the standard deviation of the predictions with both CAT and FPE models. The inset shows a detail of the first $\SI{0.5}{s}$ of plot in the top panel.
	}
	\label{fig:gw150914_intrinsic}
	\includegraphics{Images/gw150914/noise_curves.pdf}
\end{figure}

In Figs.~\ref{fig:gw150914_intrinsic}, ~\ref{fig:gw150914_extrinsic} and \ref{fig:gw150914_wf} we show the posteriors for the set of intrinsic parameters, extrinsic parameters and reconstructed waveform from our analysis. In addition, Fig.~ \ref{fig:gw150914_orders} shows the posterior distribution on the orders $p_{H}$ and $p_{L}$ for the AR processes describing the residuals in the LIGO-Hanford and LIGO-Livingston detectors. The inferred PSDs are instead shown in Fig.~\ref{fig:gw150914_psds}.

>>>>

\subsection{Marginalising over the noise PSD in GW parameter estimation: the case for GW150914}
Since the order of the AR process is not yet determined by the theory, up to this point, we opted for an in-depth investigation of several proposals in the literature and found that different loss functions are required for different situations, with the FPE loss function being the most indicated to deal with gravitational wave data. Along these lines, we directly compared the PSDs computed with MESA with the canonical Welch's algorithm. As outlined in Sec.~\ref{sec:Welch_comparison}, MESA provides PSD estimates with smaller variance and better accuracy than Welch algorithm.
The use of MESA is particularly useful for short time series samples, where Welch's method is outperformed in both precision and confidence.

Hence, as an application of MESA, and its implementation in \texttt{memspectrum}, we analyse GW150914\cite{} using a Bayesian framework that allows for the marginalisation of the order  $p$ of the AR($p$) process representing the noise data stream. The inference algorithm proceeds as usual, see \cite{}[veitch and vecchio, ashton et al, etc etc], with a modification to the likelihood calculation. The likelihood function we employ is given in Eq.~\eqref{eq:Whittle} and its calculation proceeds as follows: for all proposed Monte Carlo moves $\theta\equiv\theta_{GW}\cup p$\footnote{We indicate the set of all GW parameters (component masses, spins, luminosity distance, etc.) with $\theta_{GW}$.}, for each detector included in the analysis, we compute the detector response to the template $h(t;\theta_{GW})$ in the time-domain, subtract it from the data $d(t)$ to get the time domain residuals $r(t;\theta_{GW})$. Under the MESA assumption, we describe the PSD of the residuals as in Eq.~\eqref{eq:MESApsd}, where the order $p$ is fixed by the Monte Carlo sampler. Having computed the PSD, we Fourier transform the residuals and compute the likelihood. Our sampler is a nested sampling algorithm\cite{}[raynest] and the specific inference model is implemented as part of \texttt{granite}\cite{}[in preparation], a dedicated inference model for ground based interferometric detectors. Needless to say, despite the implementation of the fast Burg algorithm in \texttt{memspectrum}, the inference with noise marginalisation is vastly more computational expensive than the standard analysis. As our analysis template, we adopt the machine learning based MLGW model\cite{}[MLGW], an aligned spin model trained on TEOBResumS\cite{}, that has been shown to perform well on LVK detected events\cite{}[MLGW].

\begin{figure}
	\caption{Posterior distributions for $\mathcal{M}$, $q$ and spin parameters $\chi_1$ and $\chi_2$. The recovered posteriors are very similar to the LVK released ones.}
	\label{fig:gw150914_intrinsic}
	\includegraphics[width=0.5\textwidth,keepaspectratio]{Images/gw150914/intrinsic.pdf}
\end{figure}
\begin{figure}
	\caption{Posterior distributions for the sky position, (log) luminosity distance, inclination angle, polarisation angle $\psi$, reference phase $\phi_{ref}$ and geocentric coalescence time. Also for the extrinsic parameters, we see broad consistency with the LVK released ones. However, we notice that the posterior on the declination angle $\delta$ is mirrored compared to the released sky map. Nonetheless, there is substantial overlap. Similarly, the PSD marginalised posterior for the inclination angle shows a slight preference for a face-on configuration, rather than a face-off.}
	\label{fig:gw150914_extrinsic}
	\includegraphics[width=0.5\textwidth,keepaspectratio]{Images/gw150914/extrinsic.pdf}
\end{figure}
\begin{figure}
	\caption{Whitened reconstructed waveforms from our analysis for the Hanford detector (top panel) and the Livingston detector (bottom panel). The shaded turquoise area indicates the 90\% credible region.}
	\label{fig:gw150914_waveforms}
	\includegraphics[width=0.5\textwidth,keepaspectratio]{Images/gw150914/waveforms.pdf}
\end{figure}
%\begin{figure}
%	\caption{In the top panel, we show the difference between the forecasted strain values (CAT model) and the actual values observed at Livingston interferometer (blue). The red shadow denotes the $90\%$ confidence interval subtracted with the actual time series values. The confidence interval is computed on  $100$ independent random realization of the predictions.
%	The bottom panel displays the standard deviation of the predictions with both CAT and FPE models. The inset shows a detail of the first $\SI{0.5}{s}$ of plot in the top panel.
%	}
%	\label{fig:gw150914_intrinsic}
%	\includegraphics[width=0.5\textwidth,keepaspectratio]{Images/gw150914/posterior_order_H1.pdf}
%\end{figure}
%
%\begin{figure}
%	\caption{In the top panel, we show the difference between the forecasted strain values (CAT model) and the actual values observed at Livingston interferometer (blue). The red shadow denotes the $90\%$ confidence interval subtracted with the actual time series values. The confidence interval is computed on  $100$ independent random realization of the predictions.
%	The bottom panel displays the standard deviation of the predictions with both CAT and FPE models. The inset shows a detail of the first $\SI{0.5}{s}$ of plot in the top panel.
%	}
%	\label{fig:gw150914_intrinsic}
%	\includegraphics{Images/gw150914/posterior_order_L1.pdf}
%\end{figure}

\begin{figure}
	\caption{Posterior PSDs for the Hanford (orange line) and Livingstone (blue line) as inferred by our analysis.}
	\label{fig:gw150914_psd}
	\includegraphics[width=0.5\textwidth,keepaspectratio]{Images/gw150914/noise_curves.pdf}
\end{figure}

In Figs.~\ref{fig:gw150914_intrinsic}, ~\ref{fig:gw150914_extrinsic} and \ref{fig:gw150914_waveforms} we show the posteriors for the set of intrinsic parameters, extrinsic parameters and reconstructed waveform from our analysis. The inferred PSDs are instead shown in Fig.~\ref{fig:gw150914_psds}. %In addition, Fig.~ \ref{fig:gw150914_orders} shows the posterior distribution on the orders $p_{H}$ and $p_{L}$ for the AR processes describing the residuals in the LIGO-Hanford and LIGO-Livingston detectors.
